---
title: "Students-Perfomance E-Business"
output:
  pdf_document: default
  html_document:
    keep_md: true
---

# `Data Visualization And EDA`

Welcome to the Data Visualization and Exploratory Data Analysis (EDA) Notebook! This notebook is designed to help you explore and analyze your data through visualizations.

In this notebook, you will leverage various plotting techniques to create insightful visual representations of your data. By visualizing your data, you can uncover patterns, trends, and relationships that may not be immediately apparent through raw numbers alone.

Through exploratory data analysis, you will gain a deeper understanding of your data set, identify potential outliers, and explore different aspects of your data. By utilizing visualizations, you can effectively communicate your findings and draw meaningful insights from the data.

This notebook provides an efficient and organized workflow to guide you in performing data visualization and EDA. So let's get started and uncover the hidden stories within your data through captivating visualizations!

```{r}

# Install and load necessary libraries
if (!requireNamespace("caret", quietly = TRUE)) {
  install.packages("caret")
}
if (!requireNamespace("skimr", quietly = TRUE)) {
  install.packages("skimr")
}
if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}
if (!requireNamespace("janitor", quietly = TRUE)) {
  install.packages("janitor")
}
if (!requireNamespace("themis", quietly = TRUE)) {
  install.packages("themis")
}
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  install.packages("ggplot2")
}
if (!requireNamespace("corrplot", quietly = TRUE)) {
  install.packages("corrplot")
}


library(caret)
library(themis)
library(skimr)
library(tidyverse)
library(janitor)
library(dplyr)
library(ggplot2)
library(corrplot)


# Load the datasets
train <- read.csv("data/train.csv", stringsAsFactors = TRUE)
test <- read.csv("data/test.csv", stringsAsFactors = TRUE)

# Display the first few rows of the train dataset
head(train)

# Get a summary of the dataset
glimpse(train)

# Skim the dataset for a comprehensive overview
skim(train)
```

## Data Visualization

### General Distribution of Columns

The code underneath prepares a data set by converting selected columns to factor variables and creates various bar plots and histograms to visualize the distribution of different variables.

```{r}
# Convert appropriate columns to factors
test$Gender <- factor(test$Gender)
test$Job.Role <- factor(test$Job.Role)
test$Work.Life.Balance <-
  factor(test$Work.Life.Balance,
         levels = c("Poor", "Below Average", "Good", "Excellent"))
test$Job.Satisfaction <-
  factor(test$Job.Satisfaction,
         levels = c("Very Low", "Low", "Medium", "High"))
test$Performance.Rating <-
  factor(test$Performance.Rating,
         levels = c("Low", "Below Average", "Average", "High"))
test$Education.Level <-
  factor(
    test$Education.Level,
    levels = c(
      "High School",
      "Associate Degree",
      "Bachelor’s Degree",
      "Master’s Degree",
      "PhD"
    )
  )
test$Marital.Status <- factor(test$Marital.Status)

test$Job.Level <-
  factor(test$Job.Level, levels = c("Entry", "Mid", "Senior"))

test$Company.Size <-
  factor(test$Company.Size, levels = c("Small", "Medium", "Large"))

test$Remote.Work <- factor(test$Remote.Work)

test$Leadership.Opportunities <-
  factor(test$Leadership.Opportunities)

test$Innovation.Opportunities <-
  factor(test$Innovation.Opportunities)

test$Company.Reputation <-
  factor(test$Company.Reputation,
         levels = c("Very Poor", "Poor", "Good", "Excellent"))

test$Employee.Recognition <-
  factor(test$Employee.Recognition,
         levels = c("Very Low", "Low", "Medium", "High"))

# Function to create bar plots for categorical variables
create_bar_plot <- function(df, col_name, title) {
  ggplot(df, aes(x = .data[[col_name]])) +
    geom_bar(fill = "skyblue", color = "black") +
    labs(title = title,
         x = col_name,
         y = "Count") +
    theme_minimal() +
    theme(text = element_text(size = 12),
          axis.text.x = element_text(angle = 45, hjust = 1))
}

# Function to create histograms for numerical variables
create_hist_plot <- function(df, col_name, title) {
  ggplot(df, aes(x = .data[[col_name]])) +
    geom_histogram(bins = 30,
                   fill = "skyblue",
                   color = "black") +
    labs(title = title,
         x = col_name,
         y = "Count") +
    theme_minimal() +
    theme(text = element_text(size = 12))
}

# Plotting each column
plot_list <- list(
  create_hist_plot(test, "Age", "Distribution of Age"),
  create_bar_plot(test, "Gender", "Distribution of Gender"),
  create_hist_plot(test, "Years.at.Company", "Distribution of Years at Company"),
  create_hist_plot(test, "Monthly.Income", "Distribution of Monthly Income"),
  create_bar_plot(test, "Job.Role", "Distribution of Job Role"),
  create_bar_plot(test, "Work.Life.Balance", "Distribution of Work-Life Balance"),
  create_bar_plot(test, "Job.Satisfaction", "Distribution of Job Satisfaction"),
  create_bar_plot(
    test,
    "Performance.Rating",
    "Distribution of Performance Rating"
  ),
  create_hist_plot(
    test,
    "Number.of.Promotions",
    "Distribution of Number of Promotions"
  ),
  create_hist_plot(test, "Distance.from.Home", "Distribution of Distance from Home"),
  create_bar_plot(test, "Education.Level", "Distribution of Education Level"),
  create_bar_plot(test, "Marital.Status", "Distribution of Marital Status"),
  create_bar_plot(test, "Job.Level", "Distribution of Job Level"),
  create_bar_plot(test, "Company.Size", "Distribution of Company Size"),
  create_hist_plot(test, "Company.Tenure", "Distribution of Company Tenure"),
  create_bar_plot(test, "Remote.Work", "Distribution of Remote Work"),
  create_bar_plot(
    test,
    "Leadership.Opportunities",
    "Distribution of Leadership Opportunities"
  ),
  create_bar_plot(
    test,
    "Innovation.Opportunities",
    "Distribution of Innovation Opportunities"
  ),
  create_bar_plot(
    test,
    "Company.Reputation",
    "Distribution of Company Reputation"
  ),
  create_bar_plot(
    test,
    "Employee.Recognition",
    "Distribution of Employee Recognition"
  ),
  create_bar_plot(test, "Attrition", "Distribution of Attrition")
)

# Print all plots
for (p in plot_list) {
  print(p)
}

rm(plot_list)
rm(p)
rm(create_bar_plot)
rm(create_hist_plot)
```

## EDA - Exploratory Data Analysis

Exploratory Data Analysis (EDA) is an essential step in the data analysis process. It involves examining and understanding the data to uncover patterns, relationships, and potential issues. EDA utilizes descriptive statistics and data visualization techniques to gain insights and make informed decisions.

### Correlation Matrix

The correlation matrix visualizes the pairwise relationships between variables in a data set. It provides a concise summary of the strength and direction of the linear association between variables. The correlation matrix plot, created using the `corrplot` package, uses colors and shapes to represent correlation values, making it easier to identify patterns and relationships between variables. This plot helps identify variables that are highly correlated (positively or negatively) and those with weak or no correlation. It is a valuable tool for exploratory data analysis and feature selection.

```{r}
numeric_data <- test %>% mutate_all(as.numeric)


# Calculate correlation matrix
correlation_matrix <- cor(numeric_data, use = "complete.obs")

# Define color palette
col <- colorRampPalette(c("red", "grey", "blue"))(200)

# Create a correlation plot
corrplot(correlation_matrix, method = "circle", type = "upper", tl.col = "black", tl.srt = 45, col = col, tl.cex = 0.6, number.cex = 1.5)

# Convert correlation matrix to a data frame
correlation_df <- as.data.frame(as.table(correlation_matrix))

# Convert factors to characters
correlation_df$Var1 <- as.character(correlation_df$Var1)
correlation_df$Var2 <- as.character(correlation_df$Var2)


# Filter out self-correlations and duplicate pairs
correlation_df <- correlation_df %>%
  filter(Var1 != Var2) %>%
  mutate(pair = pmin(Var1, Var2) %>% paste(pmax(Var1, Var2), sep = "-")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair)


# Top 10 positive correlations
print(head(correlation_df[order(-correlation_df$Freq), ], 10))

# Top 10 negative correlations
print(head(correlation_df[order(correlation_df$Freq), ], 10))

rm(numeric_data)
rm(col)
rm(correlation_df)
rm(correlation_matrix)
```

### Work life balance by Attrition

This code calculates the count and percentage of attrition for each category of work-life balance in the "test" data set and creates a bar chart to visualize the results.

```{r}
# Calculate the count and percentage of attrition for each category of Work life balance
attrition_counts <- test %>%
  group_by(Work.Life.Balance, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Create the bar chart for Work-Life Balance with percentages
ggplot(attrition_counts, aes(x = Work.Life.Balance, y = Count, fill = factor(Attrition))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3) +
  labs(title = "Attrition by Work-Life Balance",
       x = "Work-Life Balance",
       y = "Count",
       fill = "Attrition") +
  scale_fill_manual(values = c("skyblue", "orange"),
                    labels = c("Stayed", "Left")) +
  theme_minimal()

rm(attrition_counts)
```

### Education Level by Attrition

This code calculates the count and percentage of attrition for each category of education level in the "test" data set and creates a bar chart to visualize the results.

```{r}
# Calculate the count and percentage of attrition for each category of Education Level
attrition_counts_edu <- test %>%
  group_by(Education.Level, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Create the bar chart for Education Level with percentages
ggplot(attrition_counts_edu, aes(x = Education.Level, y = Count, fill = factor(Attrition))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3, hjust = 0.5) +
  labs(title = "Attrition by Education Level",
       x = "Education Level",
       y = "Count",
       fill = "Attrition") +
  scale_fill_manual(values = c("skyblue", "orange"),
                    labels = c("Stayed", "Left")) +
  scale_x_discrete(labels = c("High School", "Associate Degree", "Bachelor's Degree", "Master's Degree", "PhD")) +
  theme_minimal()

rm(attrition_counts_edu)
```

### Job Satisfaction by Attrition

This code calculates the count and percentage of attrition for each category of job satisfaction in the "test" data set and creates a bar chart to visualize the results.

```{r}
# Calculate the count and percentage of attrition for each category of Job Satisfaction
attrition_counts_satisfaction <- test %>%
  group_by(Job.Satisfaction, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)

# Create the bar chart for Job Satisfaction with percentages
ggplot(attrition_counts_satisfaction, aes(x = Job.Satisfaction, y = Count, fill = factor(Attrition))) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5), size = 3, hjust = 0.5) +
  labs(title = "Attrition by Job Satisfaction",
       x = "Job Satisfaction",
       y = "Count",
       fill = "Attrition") +
  scale_fill_manual(values = c("skyblue", "orange"),
                    labels = c("Stayed", "Left")) +
  scale_x_discrete(labels = c("Very Low", "Low", "Medium", "High")) +
  theme_minimal()

rm(attrition_counts_satisfaction)
```

### Job Level by Attrition

This code calculates the count and percentage of attrition for each job level category in the "test" data set. It then creates a bar chart to visualize the count of attrition by job level for instances where attrition is "Left."

```{r}
attrition_counts <- test %>%
  group_by(Job.Level, Attrition) %>%
  summarise(Count = n()) %>%
  mutate(Percentage = Count / sum(Count) * 100)
# Filter the dataset to include only instances where attrition is "Left"
left_attrition <- test[test$Attrition == "Left", ]

# Create the bar chart
ggplot(left_attrition, aes(x = Job.Level)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Attrition by Job Level (Left Only)",
       x = "Job Level",
       y = "Attrition Count") +
  theme_minimal()

rm(attrition_counts)
```

### Remote Work by Attrition

This code creates a bar chart to visualize the count of attrition by remote work for instances where attrition is "Left" in the data set.

```{r}
# Filter the dataset to include only instances where attrition is "Left"
left_attrition <- test[test$Attrition == "Left", ]

# Create the bar chart
ggplot(left_attrition, aes(x = Remote.Work)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Attrition by Remote Work (Left Only)",
       x = "Remote Work",
       y = "Attrition Count") +
  theme_minimal()

rm(left_attrition)
```

### Job Level & Remote Work by Attrition

This code creates a stacked bar chart to visualize the count of attrition by job level and remote work in the data set.

```{r}
# Calculate the count of attrition for each combination of Job Level, Remote Work, and Attrition
attrition_counts <- test %>%
  group_by(Job.Level, Remote.Work, Attrition) %>%
  summarise(Count = n())

# Create the stacked bar chart with Job Level, Remote Work, and Attrition
ggplot(attrition_counts, aes(x = Job.Level, y = Count, fill = Remote.Work)) +
  geom_bar(stat = "identity") +
  labs(title = "Attrition Count by Job Level and Remote Work",
       x = "Job Level",
       y = "Count",
       fill = "Remote Work") +
  facet_wrap(~ Attrition) +
  theme_minimal()

rm(attrition_counts)
```

### Marital Status by Attrition

This code creates a bar chart to visualize the count of attrition by marital status for instances where attrition is "Left" in the data set.

```{r}
# Filter the dataset to include only instances where attrition is "Left"
left_attrition <- test[test$Attrition == "Left", ]

# Create the bar chart
ggplot(left_attrition, aes(x = Marital.Status)) +
  geom_bar(fill = "skyblue") +
  labs(title = "Attrition by Marital Status (Left Only)",
       x = "Marital Status",
       y = "Attrition Count") +
  theme_minimal()

rm(left_attrition)
```

### Distance from Home by Attrition

This code creates a bar chart to visualize the count of attrition by distance from home in the data set.

```{r}
# Convert Distance.from.Home to factor
test$Distance.from.Home <- factor(test$Distance.from.Home)

ggplot(test, aes(x = Distance.from.Home, fill = as.factor(Attrition))) +
  geom_bar() +
  labs(title = "Attrition by Distance from Home",
       x = "Distance from Home",
       y = "Count",
       fill = "Attrition") +
  theme_minimal()
```

### Number of Promotions by Attrition

This code creates a bar chart to visualize the count of attrition by the number of promotions in the data set.

```{r}
# Convert Number.of.Promotions to factor
test$Number.of.Promotions <- factor(test$Number.of.Promotions)

ggplot(test, aes(x = Number.of.Promotions, fill = as.factor(Attrition))) +
  geom_bar() +
  labs(title = "Attrition by Number of Promotions",
       x = "Number of Promotions",
       y = "Count",
       fill = "Attrition") +
  theme_minimal()
```

### Gender by Attrition

This code creates a bar chart to visualize the count of attrition by gender in the data set.

```{r}
# Convert Gender to factor
 test$Gender <- factor(test$Gender)

 ggplot(test, aes(x = Gender, fill = as.factor(Attrition))) +
   geom_bar() +
   labs(title = "Attrition by Gender",
        x = "Gender",
        y = "Count",
        fill = "Attrition") +
   theme_minimal()
```

### Age by Attrition

This code creates a box plot to compare the age distribution between different attrition statuses in the data set.

```{r}
# Box plot for Age vs Attrition
ggplot(test, aes(x = as.factor(Attrition), y = Age)) + 
  geom_boxplot(fill = "skyblue", color = "black") + 
  labs(title = "Box Plot of Age by Attrition Status",
       x = "Attrition (1 = Stayed, 2 = Left)",
       y = "Age (Years)") +
  theme_minimal()
```
